**Density Dependence**

### 2023-03-06 Rmd/report mac.Rmd
Updates to report.R (plotting), report mac.Rmd and refs.bib

@lorenzen2000allometry

@@

dropboxdir<-try(file.path(get_dropbox(), "pelagics"))

if ("try-error"%in%is(dropboxdir))
  dropboxdir="~/Dropbox/pelagics"

@@

[summary here]

@@

```{r, echo=FALSE, fig.width=10, fig.asp=0.6, fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

```{r, echo=FALSE, fig.width=10, fig.asp=0.6, fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

Natural mortality is assumed to be 0.15 for all age groups and constant over time.This value was calculated based on estimates of total mortality derived from tagging data combined with catch data (Hamre, 1980; ICES 2022; ICES 2020).

@@

Hamre, J. (1980). "Biology, exploitation, and management of the northeast atlantic mackrel." Conseil International pour l'exploration de la Mer, Rapports et Proces-verbaux des Reunions 177: 212-242.

ICES (2022). Working Group on Widely Distributed Stocks (WGWIDE 2022) Volume 4, Issue 73.

ICES (2020). Stock Annex: Mackerel (Scomber scombrus) in subareas 1-7 and 14 and divi-sions 8.a-e, 9.a (the Northeast Atlantic and adjacent waters), ICES.

```{r, echo=FALSE, fig.width=10, fig.asp=0.6, fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

NEEDS BETTER EXPLANATION [ The estimated linear model from the GAMM model relative to the observations ]

```{r, echo=FALSE, fig.width=10, fig.asp=0.6, fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

[NEEDS MORE TEXT]

A summary of the linear model is show below

```{r, echo=FALSE, fig.width=10, fig.asp=0.6, fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

*`r tab_nums("cond_gammcoefficients")`*

@@

The resulted in the parameters k=`r matPar$par[1]` and $W_{50}$=`r matPar$par[2]` which have been used in subsequent simulations.

```{r, echo=FALSE, fig.width=10, fig.asp=0.6, fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

Life history parameters are used to develop an example of modelling density dependence for mass, maturity and natural mortality-at-age. The parameters are first used to construct an `FLBRP` object representing the equilibrium, and are then coerced into an `FLStock` to model the time series dynamics.

@@

Where $\overline{W_a}$ and $\beta$ are estimated from a regression of mass-at-age ($W_a$) on total biomass ($B_a$), based on empirical data. $W_a$ is constrained so sizes do not become unfeasible large or small.

@@

Is assumed to vary at unit length ($L_1$) e.g.

$ln(M_1) =\alpha + \beta ln(L_{\infty}) + \gamma ln(k)$

where $\alpha=0.65$, $\beta=0.91$, $\gamma=0.87$

Length-at-age ($L_a$) can be obtained from the length-weight relationship

$L_a=(W/a)^{1/b}$

giving

$ln(M_1) =(W/a)^{1/b}(\alpha + \beta ln(L_{\infty}) + \gamma ln(k))$

For DD there is a relationship between Mass at age $M_a$ and total biomass ($B$)

Mackerel:

$M = 0.0232B + 0.0524$

[WHERE DO THE PARAMETERS COME FROM AND HOW TO THEY RELATE TO SUBSEQUENT ANALYSIS ??? ]

[IT IS VERY CONFUSING TO USE M OF NATURAL MORTALITY AND FOR MASS !!]

The plot below shows the equilibrium values by age for simulations initiated with different levels of biomass relative to Bmsy. The red line indicates the values from the current assessment (for natural mortality with an assumed distribution over ages). [ CHECK !]

@@

fig_nums(
  name    = "sim_dd3", level = 1, display = FALSE,
  caption = "Simulation of density dependence in mass-at-age, maturity at age and Natural mortality at age for different levels of biomass relative to Bmsy")

@@

@@

The operating models for this study are derived from the 2022 stock assessment. SAM is a statistical catch-at-age model that estimates error in the catch-at-age. In the case of mackerel, the assessment gives a higher weight to the recruitment index than to the catch data for ages 0 and 1. This means that the estimates of recruitment in the assessment can be very different from the actual number of fish that are recruited to the fishery. To ensure the MSE projections and the historical estimates are comparable, a Virtual Population Analysis (VPA) was conducted on the whole catch matrix and the survivors in the most recent year and oldest age. In the next step, we applied a natural mortality at age vector derived from the Lorenzen approach which would mimick the VPA assessment with fixed M at age. The figure below compares the SAM assessment with the VPA and the VPAM assessment. A main difference is in the recruitment estimates. Using age-varying M tends to flatten the recruitment pattern. The vpaM model was used as the base case to be compared against the density dependent simulations.

@@

fig_nums(
  name    = "om_VPADDM", level = 1, display = FALSE,
  caption = "Simulation of density dependence in Mass-at-age")

@@

fig_nums(
  name    = "om_VPADDMM", level = 1, display = FALSE,
  caption = "Simulation of density dependence in Mass-at-age and Maturity at age")

@@

fig_nums(
  name    = "om_VPADDMMM", level = 1, display = FALSE,
  caption = "Simulation of density dependence in Mass-at-age, maturity at age and M at age")

@@

The comparison of the equilibrium curves, indicate that Bmsy is expected to decrease by 30%, Fmsy is expected to increase by 113% and catch (MSY) is expected to increase by 5%. So while, inclusion of density dependent processes, has a substantial impact on biological reference points, the results change in equilibrium catch is expected to be small.

@@

**[ THIS IS CONTRADICTING THE RESULTS FROM THE EQUILIBRIUM ANALYSIS ?? ]**

>>>>

@@

dropboxdir<-try(file.path(get_dropbox(), "DDMSE"))

if ("try-error"%in%is(dropboxdir))
  dropboxdir="~/Dropbox/DDMSE"

@@

The current management of fish stocks is based on achieving the fishing mortality ( (FMSY) that achieves the maximum sustainable yield (MSY). Estimation of MSY generally only considers the stock recruitment relationship when considering density dependent population regulation mechanisms. Density dependence, however, may also operate in  other processes such as growth, sexual maturity, and natural mortality. A main aim of this study was to condition  alternatives Operating Models (OMs) for use in Management Strategy Evaluation (MSE), that allows comparison with the  assessments conducted to support the ICES advice framework. To do this the OM include density dependence in growth, maturity, natural mortality, individually and combined. We use these OMs to evaluate the consequences of not including density dependence management advice.

The Operating Models were

* Base Case, with varying M-at-age
* Density dependence in mass-at-age,
* Density dependence in mass, maturity-at-age,
* Density dependence in mass, maturity and M-at-age,

The MSY reference points BMSY and FMSY are generally based on equilibrium assumptions, either from combining yield per recruit and spawner per recruit relationship ([**Sissenwine and Shepherd, 1986**]) or by conducting long-term projections under constant fishing moralities.

Initial evaluations were conducted based on life history theory, as used by ICES to develop advice conducting MSE for data poor stocks (**Fischer et al., 2021**). **Figure 10** shows the  the equilibrium curves of catch v SSB assuming a) no density dependence, and density dependence in b) mass-at-age, c) mass & maturity-at-age, and d) mass, maturity and natural mortality-at-age. These allow the relationships between the MSY reference points and virgin biomass to be seen.

Each of the OMs were then projected for the values of FMSY based on the four assumptions about density dependence. This allows a comparison between the yield that could potentially be achieved and that forgone by making an incorrect assumption about density dependence. The OMs were first projected, without process error in recruitment, from the end of the historical estimates for a range of fishing mortality (spanning 0 to Fcrash) for 30 years into the future. The estimates in the final year of the projections were then be used to construct equilibrium curves (**Figure X2**). The points on the curve correspond to the projected outcomes for the values of FMSY from each OM. This was run as a check, as the equilibrium and projections should agree, I.e. the points at MSY should be the same colour as the curve.

The OMs were also projected for the estimates of different estimates FMSY, with stochasticty in the recruitment deviates. This allows the impact of misspecification of density dependence to be evaluated, i.e. if FMSY is underestimated then yield will be less than MSY, while if FMSY is  overestimated  yield will be lost due to overfishing. Stochastic projections of the four Operating Models for the four estimates of FMSY are shown in **Figures X3a, X3b, X3c and X3d**

**Figure X4** summaries the yields and SSBs obtained for each Operating Model under the correct and misspecified form of density dependence.

Conclusions

Estimating density-dependence from observations only possible for weight at age and maturity age. Density dependence in natural mortality can only be explored from theoretical considerations (because current assessment does not include age or time varying natural mortality).

* The impact of including density dependence: major impact on reference points. Especially Fmsy is much higher when including DD. However, the expected yields in equilibrium conditions are expected to be very similar. The main difference is in the initial step from the current Fmsy to any new Fmsy that would include DD, as the higher Fmsy would be applied to a stock that is estimated without DD. 
* The impact of ignoring density dependence
* The impact of assuming density-dependence incorrectly
* Robustness of the current ICES framework

@@

```{r, echo=FALSE, out.width = "400px", fig.align="left", message=FALSE, warning=FALSE, cache=FALSE}

@@

```{r, echo=FALSE, out.width = "400px", fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

Natural mortality is assumed to be 0.15 for all age groups and constant over time.This value was calculated based on estimates of total mortality derived from tagging data combined with catch data (@hamre1980biology; @ices2022wgwide; @ices2022macstockannex).

@@

```{r, echo=FALSE, out.width = "400px", fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

The change in relative mass-at-age (i.e. observed - mean values) is plotted against total biomass, this allows the relationship between growth and total biomass to be compared across all ages. It is seen that there is a common negative relationship for all ages. A summary of the linear model is in the table below. The trend of growth with biomass is -0.18, this was then used to model mass-at-age as:

[ CHECK THE FORMULA ]
W^'=¯W^-0.18
$W_a = {\overline{W_a}} B^{{\beta}_a}$

```{r, echo=FALSE, out.width = "400px", fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

```{r, echo=FALSE, out.width = "400px", fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

*`r tab_nums("cond_lmsummary")`*

@@

This resulted in the parameters k=`r format(matPar$par[1], digits=4)` and $W_{50}$=`r format(matPar$par[2], digits=4)` which have been used in subsequent simulations.

```{r, echo=FALSE, out.width = "400px", fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

Life history parameters are used to develop an example of modelling density dependence for mass, maturity and natural mortality-at-age. The parameters are first used to construct an *`**FLBRP** object representing the equilibrium, and are then coerced into an **FLStock** to model the time series dynamics.

@@

Where $\overline{W_a}$ and $\beta$ are estimated from a regression of mass-at-age ($W_a$) on total biomass ($B_a$), based on empirical data. $W_a$ is constrained so that sizes do not become unfeasible large or small.

@@

Natural mortality is assumed to vary at unit length ($L_1$).

Empirical and theoretical studies (@lorenzen2000allometry) have shown that natural mortality (M) decrease with fish length, but increase with the von Bertalanffy parameters of curvature k and asymptotic length L_∞. Therefore, M can be derived from life history parameters, e.g. M=1.5k(L_∞/L) (@POPE2021105952) or M=1.5k @jensen1985comparison. @lorenzen2008fish @lorenzen2022natural   showed that natural mortality rates vary with body size and age, often by orders of magnitude over the life cycle and support the existence of an allometric relationship between M and body mass (W) at-age of the form:

[Formula here]

Where M_u is M at unit weight, and d is the allometric exponent from empirical relationships shown to range from –0.3 to –0.37. d is approximately equal to –1/3, which implies that M is inversely proportional to body length, since mass is approximately proportional to the third power of length. This is consistent with [**Gislason et al., (2008)**], who proposed a relationship for natural mortality based on length

$M=\alphak(L_{\infty}χ)L\beta$

The plot below shows the equilibrium values by age for simulations initiated with different levels of biomass relative to B_MSY. The red line indicates the values from the current assessment (for natural mortality with an assumed distribution over ages). [ CHECK !]

@@

fig_nums(
  name    = "sim_dd3", level = 1, display = FALSE,
  caption = "Simulation of density dependence in mass-at-age, maturity at age and Natural mortality at age for different levels of biomass relative to Bmsy. Red line is without density dependence.")

@@

```{r, echo=FALSE, out.width = "500px",  fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

The operating models for this study are derived from the 2022 stock assessment. SAM is a statistical catch-at-age model that estimates error in the catch-at-age. In the case of mackerel, the assessment gives a higher weight to the recruitment index than to the catch data for ages 0 and 1. This means that the estimates of recruitment in the assessment can be very different from the actual number of fish that are recruited to the fishery. To ensure the MSE projections and the historical estimates are comparable, a Virtual Population Analysis (VPA) was conducted on the whole catch matrix and the survivors in the most recent year and oldest age. In the next step, we applied a natural mortality at age vector derived from the Lorenzen approach which would mimick the VPA assessment with fixed M at age. The figure below compares the SAM assessment with the VPA and the VPAM assessment. A main difference is in the recruitment estimates. Using age-varying M tends to flatten the recruitment pattern. The vpaM model was used as the base case to be compared against the density-dependent simulations.

```{r, echo=FALSE, out.width = "600px", fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@

```{r, echo=FALSE, out.width = "400px", fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

fig_nums(
  name    = "om_VPADDM", level = 1, display = FALSE,
  caption = "Simulation of density dependence in Mass-at-age. Red lines are the non-density dependent values.")

@@

```{r, echo=FALSE, out.width = "400px", fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

fig_nums(
  name    = "om_VPADDMM", level = 1, display = FALSE,
  caption = "Simulation of density dependence in Mass-at-age and Maturity at age. Red lines are the non-density dependent values.")

@@

```{r, echo=FALSE, out.width = "400px", fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

fig_nums(
  name    = "om_VPADDMMM", level = 1, display = FALSE,
  caption = "Simulation of density dependence in Mass-at-age, maturity at age and M at age. Red lines are the non-density dependent values.")

@@

The comparison of the equilibrium curves, indicate that B_MSY is expected to decrease by 30%, Fmsy is expected to increase by 113% and catch (MSY) is expected to increase by 5%. So while, inclusion of density-dependent processes, has a substantial impact on biological reference points, the resulting change in equilibrium catch is expected to be small.

```{r, echo=FALSE, out.width = "500px", fig.align="center", message=FALSE, warning=FALSE, cache=FALSE}

@@
